### 2021-05-31 papers/thesis/full/ch1-introduction/introduction.tex
Further edits, mainly including additional references.

One could say this is surprising, given that there is a long history of around 150 years associated with multivariate OPs, beginning with Hermite first presenting the multivariate Hermite polynomials in 1865 \cite{appel1926fonctions, ismail2017review}. Zernike polynomials \cite{zernike1934diffraction} were first introduced in 1934, as another example, that are a group of bivariate polynomials orthogonal on a unit circle. Koornwinder in 1975 described a method for constructing two-variable OPs from univariate OPs \cite{koornwinder1975two}. However, few books have been published over the years on the topic of multivariate orthogonal polynomials \cite{dunkl2014orthogonal}.

This branch of mathematics has an encouraging future though, not least as a basis for sparse spectral methods for solving partial differential equations (PDEs) on multidimensional domains. Spectral methods have been established on the triangle \cite{olver2019triangle} using OPs inspired by the Koornwinder approach for defining them, and on the disk \cite{vasil2016tensor} using Zernike polynomials. Applications include solving Volterra integral equations \cite{gutleb2020sparse}.

@@

Spherical Harmonics are of course a long-established and famous group of functions defined on the surface of a sphere with certain useful properties (for example, they are orthogonal to each other on the unit sphere and form a basis for expanding functions defined on the sphere) and as a result are widely used in many fields including computer graphics (e.g. \cite{moon2008efficient, sloan2013efficient}), astrophysics (e.g. \cite{vasil2019tensor}), quantum theory (e.g. \cite{varshalovich1988quantum}), geosciences (e.g. \cite{hollerbach2013parity}) and meteorology (e.g. \cite{evans1998spherical, rubinstein2015scalar}) \bstodo{cite applications of SHs}. Notably, they are also used as basis functions for the spectral transform method that makes up part of the model in the Integrated Forecasting System (IFS), which is used by ECMWF for their forecasts \cite{wedi2013fast}. While the whole sphere spectral method using the Spherical Harmonics has been successful for numerous years \cite{williamson2007evolution}, there is a drawback in the parallel scalability bottleneck that arises from the global spectral transform, which is expected to inhibit future performance of the IFS \cite{ecmwf2020scalability, wedi2013fast, diamantakisecmwf}.

@@

On the other hand, we aim to propose a strictly orthogonal \textit{polynomial} basis, and derive sparse operators for multiplication by the cartesian coordinate axes, which can in turn lead to operators for multiplication by trigonometric functions too. By utilising orthogonal polynomials, we can develop a sparse spectral method on subdomains of the sphere surface (e.g. the spherical cap) using our knowledge of other geometries and methods involving OPs developed for them.

>>>>

One could say this is surprising, given that there is a long history of around 150 years associated with multivariate OPs, beginning with Hermite first presenting the multivariate Hermite polynomials in 1865 \cite{appel1926fonctions, ismail2017review}. Zernike polynomials \cite{zernike1934diffraction} were first introduced in 1934, as another example, that are a group of bivariate polynomials orthogonal on a unit circle. Koornwinder in 1975 described a method for constructing two-variable OPs from univariate OPs \cite{koornwinder1975two}. However, few books have been published over the years on the topic of multivariate orthogonal polynomials \cite{dunkl2014orthogonal}. This branch of mathematics has an encouraging future though, not least as a basis for sparse spectral methods for solving partial differential equations (PDEs) on multidimensional domains. Spectral methods have been established on the triangle \cite{olver2019triangle} using OPs inspired by the Koornwinder approach for defining them, and on the disk \cite{vasil2016tensor} using Zernike polynomials.

@@

Spherical harmonics are of course a long-established and famous group of functions defined on the surface of a sphere with certain useful properties (for example, they are orthogonal to each other on the unit sphere and form a basis for expanding functions defined on the sphere) and as a result are widely used in many scientific fields for solving PDEs including computer graphics (e.g. \cite{moon2008efficient, sloan2013efficient}), astrophysics (e.g. \cite{vasil2019tensor}), quantum theory (e.g. \cite{varshalovich1988quantum}), biochemistry (e.g.\cite{parimal2014application, basko1998application}), geosciences (e.g. \cite{hollerbach2013parity}) and meteorology (e.g. \cite{evans1998spherical, rubinstein2015scalar, wedi2013fast, ecmwf2020scalability, diamantakisecmwf, courtier1998ecmwf}). Notably, they are also used as basis functions for the spectral transform method that makes up part of the model in the Integrated Forecasting System (IFS), which is used by ECMWF for their forecasts \cite{wedi2013fast}. While the whole sphere spectral method using the Spherical Harmonics has been successful for numerous years \cite{williamson2007evolution}, there is a drawback in the parallel scalability bottleneck that arises from the global spectral transform, which is expected to inhibit future performance of the IFS \cite{ecmwf2020scalability, wedi2013fast, diamantakisecmwf}.

@@

On the other hand, we aim to propose a strictly orthogonal \textit{polynomial} basis, and derive sparse operators for multiplication by the cartesian coordinate axes, which can in turn lead to operators for multiplication by trigonometric functions too. By utilising orthogonal polynomials in cartesian coordinates, we can develop a sparse spectral method on subdomains of the sphere surface (e.g. the spherical cap) using our knowledge of other geometries and methods involving OPs developed for them.

### 2021-05-31 papers/thesis/full/ch3-disk-slices/disk-slices.tex
Further edits, mainly including additional references.

In recent years, there has been some interesting work in the area of multivariate orthogonal polynomials in two-dimensional domains (see \cite{dunkl2014orthogonal}), such as for a Chebychev-based spectral method on the rectangle \cite{julien2009efficient} \bstodo{list some domains and cite references for them}. In particular, spectral methods have been established on the disk \cite{boyd2011comparing, vasil2016tensor} and the triangle \cite{olver2019triangle}. Jacobi polynomials are extremely useful for solving a number of problems in not just 1D but 2D also, since one can utilise the sparsity of the relations that they hold to yield operator matrices that are efficient to solve \cite{doha2006efficient}. In a similar vein to the ultraspherical method for the unit interval \cite{olver2013fast, doha2009efficient}, for the triangle a choice of OPs are the Jacobi polynomials on the triangle, which are a three-parameter family of polynomials in $(x,y)$ with differential operators acting in such a way so as to increment or decrement the parameter values \cite{dunkl2014orthogonal, koornwinder1975two, olver2019triangle}:
\bseqnnumber{
	P^{(a,b,c)}_{n,k}(x,y) := \tilde P^{(2k+b+c+1,a)}_{n-k}(x) \; (1-x)^k \; \tilde P^{(c,b)}_k\fpr({y \over 1-x}) \label{eqn:triangleops}
}
for $(x,y) \in \bsset{(x,y) \in \R^2}{0 < x < 1, \: 0 < y < 1-x}$, where $\tilde P_k^{(a,b)}$ is the degree $k$ \textit{shifted Jacobi polynomial}\footnote{The shifted Jacobi polynomials are orthogonal on the interval $[0,1]$ and given by $\tilde P_k^{(a,b)}(x) = P_k^{(a,b)}(2x - 1)$ where $P_k^{(a,b)}$ is the standard degree $k$ Jacobi polynomial.} with $a,b,c$ being parameter values.

@@

In this chapter we have shown that bivariate orthogonal polynomials can lead to sparse discretizations of general linear PDEs on specific domains whose boundary is specified by an algebraic curve -- notably here the disk-slice -- with Dirichlet boundary conditions. This work extends the triangle case \cite{beuchler2006new,li2010optimal,olver2019triangle} and whole disk case \cite{} \bstodo{cite} to non-classical geometries, and forms a building block in developing an $hp-$finite element method to solve PDEs on other polygonal domains by using suitably shaped elements (for example, by dividing the disk or a section of the disk into disk-slice elements, which has applications in turbulent pipe flow \cite{}\bstodo{cite}).

>>>>

In recent years, there has been some interesting work in the area of multivariate orthogonal polynomials in two-dimensional domains (see \cite{dunkl2014orthogonal}), such as for a Chebychev-based spectral method on the rectangle \cite{julien2009efficient} \bstodo{list some domains and cite references for them}. In particular, spectral methods have been established on the disk \cite{boyd2011comparing, vasil2016tensor} and the triangle \cite{olver2019triangle}. Jacobi polynomials are extremely useful for solving a number of problems in not just 1D but 2D also, since one can utilise the sparsity of the relations that they hold to yield operator matrices that are efficient to solve \cite{doha2006efficient}. Applications for the disk include \bstodo{give applications and references} and for the triangle include solving Volterra integral equations \cite{gutleb2020sparse}, \bstodo{give more}.

In a similar vein to the ultraspherical method for the unit interval \cite{olver2013fast, doha2009efficient}, for the triangle a choice of OPs are the Jacobi polynomials on the triangle, which are a three-parameter family of polynomials in $(x,y)$ with differential operators acting in such a way so as to increment or decrement the parameter values \cite{dunkl2014orthogonal, koornwinder1975two, olver2019triangle}:
\bseqnnumber{
	P^{(a,b,c)}_{n,k}(x,y) := \tilde P^{(2k+b+c+1,a)}_{n-k}(x) \; (1-x)^k \; \tilde P^{(c,b)}_k\fpr({y \over 1-x}) \label{eqn:triangleops}
}
for $(x,y) \in \bsset{(x,y) \in \R^2}{0 < x < 1, \: 0 < y < 1-x}$, where $\tilde P_k^{(a,b)}$ is the degree $k$ \textit{shifted Jacobi polynomial}\footnote{The shifted Jacobi polynomials are orthogonal on the interval $[0,1]$ and given by $\tilde P_k^{(a,b)}(x) = P_k^{(a,b)}(2x - 1)$ where $P_k^{(a,b)}$ is the standard degree $k$ Jacobi polynomial.} with $a,b,c$ being parameter values.

@@

In this chapter we have shown that bivariate orthogonal polynomials can lead to sparse discretizations of general linear PDEs on specific domains whose boundary is specified by an algebraic curve -- notably here the disk-slice -- with Dirichlet boundary conditions. This work extends the triangle case \cite{beuchler2006new,li2010optimal,olver2019triangle} and whole disk case \cite{vasil2016tensor} \bstodo{cite more here} to non-classical geometries, and forms a building block in developing an $hp-$finite element method to solve PDEs on other polygonal domains by using suitably shaped elements (for example, by dividing the disk or a section of the disk into disk-slice elements, which has applications in turbulent pipe flow \cite{eggels1994fully, vasil2016tensor}).
